---
title: "03 EDA"
author: "Daniel Tshiani"
date: "2026-01-02"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
working_data <- readRDS("../data/02 working data.rds")

working_data <- working_data %>%
  filter(season == 2025)
```


```{r}
glimpse(working_data)
```

```{r}
ggplot(working_data, aes(x = `gauranteed compensation`)) +
  geom_histogram(bins = 50) +
  scale_x_log10(labels = scales::dollar) +
  labs(
    title = "Distribution of MLS Players Salaries",
    x = "Salary (log scale)",
    y = "Number of Players"
  )+
  theme_minimal()
```
- Highly skewed to the right
- There are clusers. a large mass near the league minimum, a mid-tier around a few hundred thousand
- there is another peak just under the 1 million mark so that is another cluster. i am asuming its designated players
- then there are also the extrem outliers. thats the Messi's and huge players.

Even after log transformation, salaries exhibit strong tiering and extreme outliers. As a result, model performance will likely differ substantially between low- and high-salary players, so evaluation should be stratified by salary tier rather than relying on a single aggregate metric.


